---
title: "Lotka-Volterra Model of Species Competition"
author: "Andrew Tredennick and Brittany Teller"
output: html_document
runtime: shiny
---
```{r, echo=FALSE, message=FALSE}
library(shiny)
library(shinyapps)
#devtools::install_github("EcoEvoEducation/EcoEvoApps")
```
### Introduction
Here's the model

$$
\frac{dN_{1}}{dt} = r_{1}N_{1}\left(\frac{K_{1}-N_{1}-\alpha_{2,1}N_{2}}{K_{1}}\right)
$$

$$
\frac{dN_{2}}{dt} = r_{2}N_{2}\left(\frac{K_{2}-N_{2}-\alpha_{1,2}N_{1}}{K_{2}}\right)
$$

And, just in case you ever see this in a paper, it is often written in the multispecies case as:

$$
\frac{dN_{i}}{dt} = r_{i}N_{i}\left(\frac{K_{i}-N_{i}- \sum_{i \neq j}^S \alpha_{j,i}N_{j}}{K_{i}}\right)
$$

To keep everything clear, here is a table of model states and parameters.

| Parameter | Definition |
|---|---|
| $S$ | The number of susceptible (yet-to-be infected) individuals in the population | 
| $I$ | The number of infected individuals in the population |
| $R$ | The number of recovered indviduals in the population (recovered and now immune) |
| $\beta$ | Transition rate from susceptible to infected; function of disease infectiousness and individual contact rate |
| $r$ | Recovery rate |
| $\eta$ | Disease infectiousness; the probability that contact results in infection |
| $\tau$ | Individual contact rate; the rate at which individuals come into contact |


### Play with the model parameters

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(deSolve)
library(ggplot2)
library(reshape2)
library(gridExtra)
library(EcoEvoApps)
shinyAppDir(
  system.file("examples/LotkaVolterra", package="EcoEvoApps"),
  options=list(
    width="100%", height=550
  )
)
```

### Implement the model in R
